Why do we Eval?\\[2mm]\Large A large-scale study of Eval usage in R

### 2020-12-18 paper/main.tex
usage

To provide a picture of \eval usage in R, this section focuses on
library code and the \Corpus packages in our CRAN corpus. For these
packages, all runnable code in CRAN was executed, runs that did not
exercise an \eval in the corpus were discarded. The data presented
here was collected from \Nbruns runs. In total, \Allcalls invocations
of \eval and its variants were recorded. Out of the
\Staticatleastonecallsite packages that use \eval, \Triggeredpkgs
packages were observed calling it --- the difference can be chalked
down to incomplete code coverage and occasional infrastructure
failures.

@@

Figure~\ref{freq} summarizes the frequency of calls to \eval over all
runs. The left column is the number of observed calls, the right
column is the number of packages falling in that range. There are
\Fewcalls packages with low \eval frequency, i.e. fewer than 100
calls; and a similar number of packages, \Manycalls to be precise, use
\eval more than 1,000 times. One package, \Maxcallspack, calls it
\Maxcalls times and thus accounts for over half of the observed
invocations of \eval.

@@

Figure~\ref{tab:variantseval} illustrates the use of variants of
\eval. For each of the four variants, the first rows shows the number
of call sites occurring in the corpus (\emph{static)}), the second row
is the number of call sites that were encountered during tracing
(\emph{exercised}), the last row is the number of calls
(\emph{invocations}). The vast majority of call sites, and calls, are
to \eval itself; \c{evalq} and \c{local} are barely used at all.

\begin{wrapfigure}{r}{5.8cm}
  \begin{tabular}{|r@{\,}r@{\,}l@{\,}r|r@{\,}r@{\,}l@{}r|} \hline
\multicolumn{3}{|c}{\small\#calls} &\small\#sites &
\multicolumn{3}{c}{\small\#calls} &\small\#sites \\\hline
\tt 0 &--& \tt 50    & \Runbina & \tt 501 &--& \tt 1000   & \Runbine\\
\tt 51 &--& \tt 100  & \Runbinb & \tt 1001 &--& \tt 1500  & \Runbinf\\
\tt 101 &--& \tt 250 & \Runbinc & \tt 1501 &--& \tt 2000  & \Runbing\\
\tt 251 &--& \tt 500 & \Runbind & \tt 2001 &--& \tt 3000 & \Runbinh\\\hline
\end{tabular}
\caption{Normalized invocations} \label{tab:callsitesnormalized}
\end{wrapfigure}

Figure~\ref{tab:callsitesnormalized} shows normalized invocation
counts. The column on the left is the average number of invocations of
\eval from a particular call site for any given run. The number on the
right is number of call sites that fall in that range. For instance,
there are \Runbinh call sites that are invoked on average over 2,001
times in any run they are exercised.

\medskip

\begin{wrapfigure}{r}{5.5cm}
  \vspace{-8mm}
    \includegraphics[width=.4\textwidth]{calls_per_run_per_call_site}
    \caption{Normalized invocations
      (small)} \label{fig:callsitesnormalized}
   \vspace{-8mm}
\end{wrapfigure}

The majority of call sites, \Runbina, are exercised up to 50 times.
Larger numbers of calls suggest that \eval may be called within a loop
or from a recursive context -- this seems to be the exception.
Figure~\ref{fig:callsitesnormalized} zooms in on call sites with an
average of fewer than 51 invocation. The x-axis shows the number of
invocations and the y-axis is the number of call sites. Most call
sites are invoked once, about half as many are invoked twice. After
that the frequency quickly drops.

Analysis of the arguments passed to \eval reveals that XXX are of kind
\c{symbol} (single variables), XXX are \c{language} (an internal R
type roughly corresponding to a single expression), and XXX are
\c{expression} (one or more statements). The remaining arguments are R
values which are returned unchanged.

\mypara{Amount of code loaded by eval.} We use \c{deparse} to turn the
input to \eval into strings and measure the size of that input. The
largest input is \Maxszeval and the median input size is
\Medianszeval. Most inputs are small, less than \SzninefiveRnd
bytes. The average size is \Avgszeval bytes.

\mypara{Amount of computation via eval.} We compare the number of
events intercepted during execution of \eval across the whole corpus.
For this, we use the number of interpreter steps as a proxy for
events. We observe \AllEventCountRnd \NOTE{Check that number.}
events all told, and \EventsRnd events occur inside eval
(\EvalEventAllPerc of all events). Figure~\ref{fig:eventsPerPackage}
show a violin plot of events.

\begin{figure}[h]
	\centering
	\includegraphics[width=.4\textwidth]{events_per_pack_small}
	\includegraphics[width=.4\textwidth]{events_per_pack_large}
	\caption{Events per packages} \label{fig:eventsPerPackage}
\end{figure}

We observe a very wide spread in the number of events generated by
\eval calls. The largest number of events generated by an \eval call
is \MaxeventsRnd. However, a majority (\Smalleventspct) of \eval calls
performs up to 50 events suggesting that most expressions passed to
\eval are small. It is interesting to note that a higher proportion of
Core R \eval calls perform side effects in the
\EventsMinRangeB--\EventsMaxRangeB range. Only \EventsCoreEvalCountC
Core R \eval calls and \EventsPackageEvalCountC package \eval call
generate \EventsMinRangeC to \EventsMaxRangeC events. These \eval
calls originate from statistical modeling packages such as \mlogit,
\mboost, \metafor, \lavaan, \mclust and \gamlss.

\NOTE{Talk about the percentage of code executed in eval}

>>>>

To provide a picture of \eval usage in R, this section focuses on
library code in the \Corpus packages of our CRAN corpus. For these,
all runnable code in CRAN was executed, runs that did not exercise an
\eval from the corpus were discarded; \Nbruns runs were retained and
generated the data presented here. In total, \Allcalls invocations of
\eval and its variants were recorded. Out of the
\Staticatleastonecallsite packages that use \eval, \Triggeredpkgs
packages were observed calling it --- the difference can be chalked
down to incomplete code coverage and occasional infrastructure
failures.

@@

Figure~\ref{freq} summarizes the frequency of calls to \eval over all
runs. The left column is the number of observed calls, the right
column is the number of packages falling in that range. There are
\Fewcalls packages with low \eval frequency, i.e. fewer than 100
calls; and a similar number of packages, \Manycalls to be precise,
that use \eval more than 1,000 times. One package, \Maxcallspack,
calls it \Maxcalls times and thus accounts for over half of the
observed invocations of \eval.

@@

Figure~\ref{tab:variantseval} illustrates the use of variants of
\eval. For each of the four variants, the first rows shows the number
of call sites occurring in the corpus (\emph{static)}), the second row
is the number of call sites that were encountered during tracing
(\emph{exercised}), the last row is the number of calls
(\emph{invocations}). The vast majority of call sites, and calls, are
to \eval itself; \c{evalq} and \c{local} are barely used at all. The
difference between static and exercised reflects spotty code coverage.

\begin{wrapfigure}{r}{5.8cm}
  \begin{tabular}{|r@{\,}r@{\,}l@{\,}r|r@{\,}r@{\,}l@{}r|} \hline
\multicolumn{3}{|c}{\small\#calls} &\small\#sites &
\multicolumn{3}{c}{\small\#calls} &\small\#sites \\\hline
\tt 0 &--& \tt 50    & \Runbina & \tt 501 &--& \tt 1000   & \Runbine\\
\tt 51 &--& \tt 100  & \Runbinb & \tt 1001 &--& \tt 1500  & \Runbinf\\
\tt 101 &--& \tt 250 & \Runbinc & \tt 1501 &--& \tt 2000  & \Runbing\\
\tt 251 &--& \tt 500 & \Runbind & \tt 2001 &--& \tt 3000 & \Runbinh\\\hline
\end{tabular}
\caption{Normalized invocations} \label{tab:callsitesnormalized}

\includegraphics[width=.4\textwidth]{calls_per_run_per_call_site}
    \caption{Normalized invocations
      (small)} \label{fig:callsitesnormalized}
   \vspace{-8mm}
\end{wrapfigure}

Figure~\ref{tab:callsitesnormalized} shows normalized invocation
counts. The column on the left is the average number of invocations of
\eval from a particular call site for any given run. The number on the
right is number of call sites that fall in that range. For instance,
there are \Runbinh call sites that are invoked on average over 2,001
times in any run they are exercised.
Larger numbers of calls suggest that \eval may be called within a loop
or from a recursive context -- this seems to be the exception as the
majority of call sites, \Runbina, are exercised less than 51 times.
Figure~\ref{fig:callsitesnormalized} zooms in on low frequency call
sites. The x-axis shows the normalized invocations and the y-axis is
the number of call sites for that value. The figure shows that most
call sites are invoked only once, about half as many are invoked
twice. After that the frequency of invocation quickly drops.

Analysis of the arguments passed to \eval reveals that \Symbolpercent
are of kind \c{symbol} (these are single variables such as \c{x}),
\Languagepercent are \c{language} (an internal R type roughly
corresponding to a single expression, such as \c{x+1}), and
\Expressionpercent are \c{expression} (one or more statements). These
account for \Codepercent of the invocations, the arguments to the
remaining invocations are R values which are returned unchanged by
\eval.

To estimate how much executable code is injected into the program
through \eval, we measure the size of arguments in number of
characters. This measurement is obtained by de-parsing every argument
to \eval. Our de-narser does not emit code for values as some values
can be very large and would overwhelm the code. On average, arguments
to \eval are \Avgszeval characters long, the median size is
\Medianszeval. The large input observed is \Maxszeval.
Figure~\ref{fig:sizedistribution} shows the distribution of sizes. The
x-axis is the size of arguments in characters, and the y-axis is the
count of arguments with that size.

\begin{figure}[h!]
 \includegraphics[width=\textwidth]{size_loaded_distribution}
\caption{Loaded code} \label{fig:sizedistribution}
\end{figure}

To measure the work performed in \evals, we instrument the R
interpreter to count the number of operations performed during
evaluation of the argument. For the purposes of this experiment, each
interpreter step is an \emph{event}. For our corpus, \EvalEventAllPerc
of events (\EventsRnd) are performed within \eval. \Smalleventspct of
\evals perform 50 or fewer events. Figure~\ref{fig:eventsPerPackage}
shows two violin plots. The plot on the left is for evals with fewer
or equal than 50 events, the plot on the right is for more
work-intensive \evals. The biggest \eval performs \MaxeventsRnd
events.

\begin{figure}[h!]
	\centering
	\includegraphics[width=.4\textwidth]{events_per_pack_small}
	\includegraphics[width=.4\textwidth]{events_per_pack_large}
	\caption{Events per packages} \label{fig:eventsPerPackage}
\end{figure}
